---
title: "R Notebook"
output: 
  html_notebook: 
    toc: yes
---

# 06/04/2017, Thursday

## Read CMap data

```{r}
filesx=c('scripts','results','data','docs')
sapply(paste('mkdir',filesx),system)
```

Save this file under scripts.

```{r}
system('mkdir -p data/raw/CMap/')
file.copy(from = '~/GD/ageingDrugs_cMAP/data/raw/CMap/rankMatrix.txt',to = 'data/raw/CMap/',copy.date = T)
```

```{r}
library(data.table)
rankmatrix=fread('data/raw/CMap/rankMatrix.txt',skip = 1)
```

```{r}
colnames(rankmatrix)=strsplit(readLines('data/raw/CMap/rankMatrix.txt',n = 1),'\t')[[1]]
rankmatrix[1:10,1:10]
```

```{r}
file.copy(from = '~/GD/ageingDrugs_cMAP/data/raw/CMap/cmap_instances_02.csv',to = 'data/raw/CMap/',copy.date = T)
```

```{r}
instances=read.csv('data/raw/CMap/cmap_instances_02.csv')
instances[1:10,1:10]
```

```{r}
system('mkdir -p data/processed/CMap')
save(rankmatrix,instances,file='data/processed/CMap/rank_and_instances.RData')
rm(list=ls())
```


# 10/04/2017, Monday

## Construct the correlation and p matrices

Copy the age-series gene expression change results and ages under /data/processed/humanBrainMicroarray/exp_cors

```{r}
system('mkdir -p ./data/processed/humanBrainMicroarray/exp_cors')
sapply(paste('cp',
             list.files('~/GD/ageingDrugs_cMAP-ca784cc95f522a5cf2ce687fed49f1c4ae1493d2/preliminary2/data/processed/humanBrain/humanBrain_expCors/',
                        full.names = T),
             './data/processed/humanBrainMicroarray/exp_cors/'),
       function(x)system(x))
```

```{r}
filesx=list.files('./data/processed/humanBrainMicroarray/exp_cors/',full.names = T)
filesx=filesx[!grepl('Lu2004',filesx)]
filesx=grep('age.rds',filesx,v=T)
cors=lapply(filesx,readRDS)
nms=gsub('_age.rds','',sapply(strsplit(filesx,'/'),function(x)x[7]))
names(cors)=nms
genelist=unique(unname(unlist(sapply(cors,rownames))))
cormat=matrix(NA,nrow=length(genelist),ncol=length(nms),dimnames = list(genelist,nms))
pmat=matrix(NA,nrow=length(genelist),ncol=length(nms),dimnames = list(genelist,nms))
padjmat=matrix(NA,nrow=length(genelist),ncol=length(nms),dimnames = list(genelist,nms))
for(nm in nms){
  cor=cors[[nm]]
  genes=rownames(cor)
  cormat[genes,nm]=cor[,1]
  pmat[genes,nm]=cor[,2]
  padjmat[genes,nm]=cor[,3]
}
print(cormat[1:10,1:5])
print(pmat[1:10,1:5])
print(padjmat[1:10,1:5])
saveRDS(cormat,file='./data/processed/humanBrainMicroarray/cormat.rds')
saveRDS(pmat,file='./data/processed/humanBrainMicroarray/pmat.rds')
saveRDS(padjmat,file='./data/processed/humanBrainMicroarray/padjmat.rds')
rm(list=ls())
```

## Basic statistics on datasets:

Read the data.
```{r}
cormat=readRDS('./data/processed/humanBrainMicroarray/cormat.rds')
pmat=readRDS('./data/processed/humanBrainMicroarray/pmat.rds')
padjmat=readRDS('./data/processed/humanBrainMicroarray/padjmat.rds')
```


Number of sub-datasets:
```{r}
ncol(cormat)
```


Number of datasources:
```{r}
nms=colnames(cormat)
length(unique(sapply(strsplit(nms,'_'),function(x)x[1])))
```


Number of genes detected in total:
```{r}
nrow(cormat)
```


```{r}
library(knitr)
dsdist=data.frame(`Number_of_Genes`=colSums(!is.na(cormat)))
dsdist$Dataset=rownames(dsdist)
ggplot(dsdist,aes(x=Dataset,y=Number_of_Genes))+
  geom_bar(stat='identity')+
  theme_bw()+
  theme(axis.text.x = element_text(angle=90))+
  ylab('Number of genes')+ 
  xlab('')+
  ggtitle('Number of genes detected in each of the sub-datasets')
```


```{r}
library(reshape2)
library(ggplot2)
genedist=melt(rowSums(!is.na(cormat)))
ggplot(genedist,aes(x=value))+
  geom_bar()+
  theme_bw()+
  ylab('Number of genes')+ 
  xlab('Number of datasets')+
  ggtitle('How many genes are detected in how many datasets?')
```

```{r,fig.height=4,fig.width=4}
signs=data.frame(genes=rownames(cormat),
           increase=rowSums(sign(cormat)==1,na.rm=T),
           decrease=rowSums(sign(cormat)==-1,na.rm=T))
signs$total=signs$increase+signs$decrease
signs$consistent=factor(as.numeric(apply(signs,1,function(x)max(x[2],x[3]))),levels=25:1)
head(signs)
signs$color=signs$total==25 & as.numeric(as.character(signs$consistent))==25
ggplot(signs,aes(x=consistent,fill=color))+
  geom_bar(stat='count')+
  facet_wrap(~total,scales='free_x')+
  theme_bw()+
  xlab('Number of Consistent Changes')+
  ggtitle('How many consistent changes?')+
  guides(fill=F)+
  scale_fill_manual(values=c('gray30','darkred'))+
  theme(axis.text.x = element_text(angle=90))
```

## Consistent Changes
```{r}
cons_up=rownames(cormat)[which(rowMeans(cormat>0)==1)]
cons_down=rownames(cormat)[which(rowMeans(cormat<0)==1)]
bg=names(which(rowMeans(is.na(cormat))==0))
kable(data.frame(Up=length(cons_up),Down=length(cons_down),Background=length(bg)))
```

## GO Enrichment:

```{r}
library(clusterProfiler)
go_up_bp=enrichGO(gene = cons_up,
               universe = bg,
               keytype = 'ENSEMBL',
               OrgDb=org.Hs.eg.db,
               ont='BP',
               minGSSize = 10,
               readable = T)

go_down_bp=enrichGO(gene = cons_down,
               universe = bg,
               keytype = 'ENSEMBL',
               OrgDb=org.Hs.eg.db,
               ont='BP',
               minGSSize = 10,
               readable = T)

go_change_bp=enrichGO(gene = c(cons_up,cons_down),
               universe = bg,
               keytype = 'ENSEMBL',
               OrgDb=org.Hs.eg.db,
               ont='BP',
               minGSSize = 10,
               readable = T)

go_up_mf=enrichGO(gene = cons_up,
               universe = bg,
               keytype = 'ENSEMBL',
               OrgDb=org.Hs.eg.db,
               ont='MF',
               minGSSize = 10,
               readable = T)

go_down_mf=enrichGO(gene = cons_down,
               universe = bg,
               keytype = 'ENSEMBL',
               OrgDb=org.Hs.eg.db,
               ont='MF',
               minGSSize = 10,
               readable = T)

go_change_mf=enrichGO(gene = c(cons_up,cons_down),
               universe = bg,
               keytype = 'ENSEMBL',
               OrgDb=org.Hs.eg.db,
               ont='MF',
               minGSSize = 10,
               readable = T)


go_up_cc=enrichGO(gene = cons_up,
               universe = bg,
               keytype = 'ENSEMBL',
               OrgDb=org.Hs.eg.db,
               ont='cc',
               minGSSize = 10,
               readable = T)

go_down_cc=enrichGO(gene = cons_down,
               universe = bg,
               keytype = 'ENSEMBL',
               OrgDb=org.Hs.eg.db,
               ont='cc',
               minGSSize = 10,
               readable = T)

go_change_cc=enrichGO(gene = c(cons_up,cons_down),
               universe = bg,
               keytype = 'ENSEMBL',
               OrgDb=org.Hs.eg.db,
               ont='cc',
               minGSSize = 10,
               readable = T)
```

```{r}
plot_mygores=function(myres){
  library(igraph)
  library(GGally)
  mylist=setNames(strsplit(myres@result$geneID,'/'),myres@result$Description)
  mylist=melt(mylist)
  gos=mylist$L1
  genes=mylist$value
  mygraph=graph_from_data_frame(mylist, directed = F)
  cx=1+names(V(mygraph))%in%unique(gos)
  sx=setNames(as.numeric(sapply(strsplit(myres@result$BgRatio,'/'),function(x)x[1])),myres@result$Description)
  sizesx=cx*3
  sizesx[sizesx==6]=log2(sx)
  ggnet2(mygraph,
         color = cx,
         label = T,
         label.size = c(1.5,3)[cx],
         size = sizesx,
         max_size = 12,
         palette = 'Dark2')+
    guides(size=F,color=F)
}
```

Save the graphs:

```{r}
system('mkdir -p ./results/humanBrainMicroarray/GO')
pdf('./results/humanBrainMicroarray/GO/go_up_bp.pdf')
plot_mygores(go_up_bp)
dev.off()
pdf('./results/humanBrainMicroarray/GO/go_up_mf.pdf')
plot_mygores(go_up_mf)
dev.off()
pdf('./results/humanBrainMicroarray/GO/go_down_bp.pdf',width=10,height = 10)
plot_mygores(go_down_bp)
dev.off()
pdf('./results/humanBrainMicroarray/GO/go_down_mf.pdf',width=10,height = 10)
plot_mygores(go_down_mf)
dev.off()
pdf('./results/humanBrainMicroarray/GO/go_down_cc.pdf',width=10,height = 10)
plot_mygores(go_down_cc)
dev.off()
pdf('./results/humanBrainMicroarray/GO/go_change_bp.pdf',width=15,height = 15)
plot_mygores(go_change_bp)
dev.off()
pdf('./results/humanBrainMicroarray/GO/go_change_cc.pdf',width=15,height = 15)
plot_mygores(go_change_cc)
dev.off()
```

### Up vs Background

```{r}
as.data.frame(go_up_bp)
```


```{r}
plotGOgraph(go_up_bp,nrow(as.data.frame(go_up_bp)))
```

```{r}
as.data.frame(go_up_mf)
```

```{r}
plotGOgraph(go_up_mf,nrow(as.data.frame(go_up_mf)))
```

### Down vs Background

```{r}
as.data.frame(go_down_bp)
```


```{r}
plotGOgraph(go_down_bp,nrow(as.data.frame(go_down_bp)))
```

```{r}
as.data.frame(go_down_mf)
```

```{r}
plotGOgraph(go_down_mf,nrow(as.data.frame(go_down_mf)))
```

```{r}
as.data.frame(go_down_cc)
```


```{r}
plotGOgraph(go_down_cc,nrow(as.data.frame(go_down_cc)))
```

### All Consistent Change vs Background

```{r}
as.data.frame(go_change_bp)
```


```{r}
plotGOgraph(go_change_bp,nrow(as.data.frame(go_change_bp)))
```

```{r}
as.data.frame(go_change_cc)
```


```{r}
plotGOgraph(go_change_cc,nrow(as.data.frame(go_change_cc)))
```

## KEGG Enrichment

```{r}
signs$sign=c(-1,1)[1+(signs$increase>signs$decrease)]
signs$frac=(as.numeric(as.character(signs$consistent))/25)*signs$sign
genelist=setNames(signs$frac,signs$genes)
genelist=genelist[bg]
```

```{r}
eg = bitr(names(genelist), fromType="ENSEMBL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
bg_e = unique(eg$ENTREZID[eg$ENSEMBL%in%names(genelist)])
cons_up_e = unique(eg$ENTREZID[eg$ENSEMBL%in%cons_up])
cons_down_e = unique(eg$ENTREZID[eg$ENSEMBL%in%cons_down])
keg_up=enrichKEGG(gene = cons_up_e,universe = bg_e,pvalueCutoff = 1,pAdjustMethod = 'BH',organism = 'hsa',minGSSize = 10,qvalueCutoff = 0.1)
keg_down=enrichKEGG(gene = cons_down_e,universe = bg_e,pvalueCutoff = 1,pAdjustMethod = 'BH',organism = 'hsa',minGSSize = 10,qvalueCutoff = 0.1)
keg_anychange=enrichKEGG(gene = c(cons_up_e,cons_down_e),universe = bg_e,pvalueCutoff = 1,pAdjustMethod = 'BH',organism = 'hsa',minGSSize = 10,qvalueCutoff = 0.1)
genelist_e=sapply(bg_e,function(x){
  ensver=eg$ENSEMBL[eg$ENTREZID==x]
  if(length(ensver)==1){unname(genelist[names(genelist)%in%ensver])}
  else {NA}
})
library("pathview")
system('mkdir -p ./results/humanBrainMicroarray/KEGG')
sapply(keg_up@result$ID,function(path){
  pathview(gene.data  = genelist_e,pathway.id = path,species= "hsa",kegg.dir = './results/humanBrainMicroarray/KEGG')})
sapply(keg_down@result$ID,function(path){
  pathview(gene.data  = genelist_e,pathway.id = path,species= "hsa",kegg.dir = './results/humanBrainMicroarray/KEGG')})
sapply(keg_anychange@result$ID,function(path){
  pathview(gene.data  = genelist_e,pathway.id = path,species= "hsa",kegg.dir = './results/humanBrainMicroarray/KEGG')})
```

```{r}
as.data.frame(keg_up)
as.data.frame(keg_down)
as.data.frame(keg_anychange)
```

## Connectivity Map analysis

### Convert IDs
```{r}
idconvert=readRDS('~/GD/ageingDrugs_cMAP-ca784cc95f522a5cf2ce687fed49f1c4ae1493d2/preliminary2/data/processed/humanBrain/rdss/id_convert.rds')
cons_up_affy=setdiff(unique(idconvert$affy_hg_u133a[idconvert$ensembl_gene_id%in%cons_up]),c(NA,''))
cons_down_affy=setdiff(unique(idconvert$affy_hg_u133a[idconvert$ensembl_gene_id%in%cons_down]),c(NA,''))
write.table(cons_up_affy,file='./data/processed/humanBrainMicroarray/consistent_up.grp',row.names=F,col.names=F,quote = F)
write.table(cons_down_affy,file='./data/processed/humanBrainMicroarray/consistent_down.grp',row.names=F,col.names=F,quote = F)
```

### Load results

```{r}
cmap=read.csv('./data/processed/humanBrainMicroarray/CMap_results/20170410_consistent_byname.csv')
cmap_cellline=read.csv('./data/processed/humanBrainMicroarray/CMap_results/20170410_consistent_bynameandcellline.csv')
p=as.numeric(as.character(cmap$p))
p[complete.cases(p)]=p.adjust(p[complete.cases(p)],method='fdr')
cmap$padj=p
p=as.numeric(as.character(cmap_cellline$p))
p[complete.cases(p)]=p.adjust(p[complete.cases(p)],method='fdr')
cmap_cellline$padj=p
cmap[which(cmap$padj<0.1),]
cmap_cellline[which(cmap_cellline$padj<0.05),]
```

```{r}
save(list=ls(),file='data/processed/humanBrainMicroarray/analysis.RData')
```

