---
title: "Plus and Minus Drugs - Genes"
output:
  html_notebook: default
  html_document: 
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

# Data preparation

```{r,message=F,warning=F}
library(biomaRt)
library(tidyverse)
library(data.table)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(BSDA)
```

```{r,warning=F,message=F}
array_summary=read_csv('./data/processed/humanBrainMicroarray/CMap_results/all_consistent_byname.csv')
array_detailed=read_csv('./data/processed/humanBrainMicroarray/CMap_results/all_consistent_detailed.csv')
gtex_summary=read_csv('./data/processed/brainGTEx/CMap_results/GTEx_brain.csv')
gtex_detailed=read_csv('./data/processed/brainGTEx/CMap_results/gtex_brain_detailed.csv')
```

```{r,message=F,warning=F}
array_summary$p.adj=p.adjust(as.numeric(array_summary$p),method='fdr')
gtex_summary$p.adj=p.adjust(as.numeric(gtex_summary$p),method='fdr')
```

```{r}
array=full_join(array_summary,array_detailed,by='cmap name')
colnames(array)=c('DrugRank','Name','MeanScore','n','Enrichment','p','Specificity','Non-NULL','Adj_p','ExperimentRank','Batch','Dose','CellLine','Score','Up','Down','InstanceID')
gtex=full_join(gtex_summary,gtex_detailed,by='cmap name')
colnames(gtex)=c('DrugRank','Name','MeanScore','n','Enrichment','p','Specificity','Non-NULL','Adj_p','ExperimentRank','Batch','Dose','CellLine','Score','Up','Down','InstanceID')
combined=full_join(array,gtex,by='InstanceID')
head(combined)
combined=combined[,-c(19,21,28,29,30)]
colnames(combined)=c('Array_DrugRank','Name','Array_MeanScore','n','Array_Enrichment','Array_p','Array_Specificity','Array_Non-NULL','Array_ADJp',
                     'Array_ExperimentRank','Batch','Dose','CellLine','Array_Score','Array_Up','Array_Down','InstanceID',
                     'GTEx_DrugRank','GTEx_MeanScore','GTEx_Enrichment','GTEx_p','GTEx_Specificity','GTEx_Non-NULL','GTEx_ADJp',
                     'GTEx_ExperimentRank','GTEx_Score','GTEx_Up','GTEx_Down')
rm(list=setdiff(ls(),'combined'))
```

```{r}
head(combined)
```

```{r,message=F,warning=F}
array_down=readRDS('./data/processed/humanBrainMicroarray/all_down_affy.rds')
array_up=readRDS('./data/processed/humanBrainMicroarray/all_up_affy.rds')
gtex_down=readLines('./data/processed/brainGTEx/down_affy.grp')
gtex_up=readLines('./data/processed/brainGTEx/up_affy.grp')
drugMat=fread('data/raw/CMap/amplitudeMatrix.txt')
drug_colx=read.table('data/raw/CMap/amplitudeMatrix.txt',nrows = 1)
```

```{r}
colnames(drugMat)=as.character(drug_colx[,as.character(colnames(drugMat))])
```

```{r}
drugMat=drugMat[which(drugMat$`1`%in%c(array_up,array_down)),]
```


```{r,message=F,warning=F}
drugTable=melt(drugMat)
colnames(drugTable)=c('ProbeID','InstanceID','DrugChange')
head(drugTable)
```

```{r}
drugTable$ArrayChange='-'
drugTable$ArrayChange[drugTable$ProbeID%in%array_up]='Up'
drugTable$ArrayChange[drugTable$ProbeID%in%array_down]='Down'

drugTable$GTExChange='-'
drugTable$GTExChange[drugTable$ProbeID%in%gtex_up]='Up'
drugTable$GTExChange[drugTable$ProbeID%in%gtex_down]='Down'
```

```{r}
rm(list=setdiff(ls(),c('combined','drugTable')))
drugTable$InstanceID=as.character(drugTable$InstanceID)
combined$InstanceID=as.character(combined$InstanceID)
```

```{r}
final_combined=full_join(combined,drugTable,by='InstanceID')
```

```{r}
head(final_combined)
```

```{r}
rm(list=setdiff(ls(),'final_combined'))
```

```{r}
signiflist=filter(final_combined,Array_ADJp<=0.05 | GTEx_ADJp<=0.05)
rm('final_combined')
head(signiflist)
dim(signiflist)
```

# Analysis of the Array Genes

```{r,fig.width=16,fig.height=2.5}
ggplot(signiflist,aes(x=DrugChange,fill=(Array_MeanScore)>0))+
  facet_grid(ArrayChange~Name,scales='free')+
  geom_density()+
  theme_bw()+
  guides(fill=guide_legend('(+) Drug'))
```

```{r}
xx=signiflist %>% 
  mutate(ArrayPositive=Array_MeanScore>0)%>%
  group_by(ProbeID,ArrayPositive) %>%
  select(Name,Array_MeanScore,Array_Score,Array_Down,Array_Up,DrugChange,ArrayChange) %>%
  unique() %>%
  group_by(ProbeID,Name) %>%
  summarise(percpos=mean(DrugChange>0)) %>%
  spread(value = percpos,key = ProbeID) %>%
  as.data.frame()

rownames(xx)=xx$Name
xx$Name=NULL
xx=as.matrix(xx)
```

```{r}
xx[1:5,1:5]
```


```{r}
colannot=signiflist %>%
  select(ProbeID,ArrayChange) %>%
  unique()
colannot=as.data.frame(colannot)
rownames(colannot)=colannot$ProbeID
colannot$ProbeID=NULL
```

## All genes - heatmap

```{r}
pheatmap(t(xx),
         annotation_row = colannot,
         color = colorRampPalette(c('blue','gray95','red'))(19),
         breaks = seq(0,abs(max(xx)),length.out = 20),
         show_rownames = F,
         # kmeans_k = 50,
         cutree_rows = 4,cutree_cols = 4)
```

```{r,message=F,warning=F}
drugInfo=read_csv('./druglist/drugInfo_final3.csv')
names(drugInfo)[1]='Name'
drugInfo$Name=unname(setNames(sort(unique(signiflist$Name)),sort(unique(drugInfo$Name)))[drugInfo$Name])
```

```{r}
combined=full_join(signiflist,drugInfo,'Name')
combined$DrugAge=rowSums((combined%>%select(`Rat Avg. Lifespan Extension`:`Yeast Avg. Lifespan Extension`))!='-')>0
combined
```

```{r}
# saveRDS(combined,file='./data/processed/humanBrainMicroarray/drugProfiles_combined.rds')
rm(list=ls())
```


## Investigate causes and responses in ageing

```{r}
combined=readRDS('./data/processed/humanBrainMicroarray/drugProfiles_combined.rds')
xx=combined %>%
  filter(DrugAge==T) %>%
  group_by(Name,ProbeID) %>%
  summarise(percPos=mean(DrugChange>0),
            percNeg=mean(DrugChange<0),
            mean=mean(DrugChange),
            median=median(DrugChange)) %>%
  ungroup() %>%
  unique() %>%
  group_by(ProbeID) %>%
  summarise(percPos_75higher=mean(percPos>0.6),
            percNeg_75higher=mean(percNeg>0.6),
            perc_meanPos=mean(mean>0),
            perc_meanNeg=mean(mean<0),
            perc_medianPos=mean(median>0),
            perc_medianNeg=mean(median<0))
drugPos=(xx %>% filter(percPos_75higher>=0.5) %>% select(ProbeID))$ProbeID
drugNeg=(xx %>% filter(percNeg_75higher>=0.5) %>% select(ProbeID))$ProbeID

resx=combined %>% 
  select(ProbeID,ArrayChange) %>% 
  unique()

resx$ProLong='-'
resx$ProLong[resx$ProbeID%in%drugPos]='Up'
resx$ProLong[resx$ProbeID%in%drugNeg]='Down'
DU=filter(resx,ArrayChange=='Down' & ProLong=='Up')$ProbeID
UD=filter(resx,ArrayChange=='Up' & ProLong=='Down')$ProbeID
UU=filter(resx,ArrayChange=='Up' & ProLong=='Up')$ProbeID
DD=filter(resx,ArrayChange=='Down' & ProLong=='Down')$ProbeID
table(resx$ArrayChange,resx$ProLong)
```

```{r}
martx=useMart('ensembl','hsapiens_gene_ensembl')
DU_E=unique(getBM(attributes = 'ensembl_gene_id',filters = 'affy_hg_u133a',values = DU,mart = martx)[,1])
UD_E=unique(getBM(attributes = 'ensembl_gene_id',filters = 'affy_hg_u133a',values = UD,mart = martx)[,1])
DD_E=unique(getBM(attributes = 'ensembl_gene_id',filters = 'affy_hg_u133a',values = DD,mart = martx)[,1])
UU_E=unique(getBM(attributes = 'ensembl_gene_id',filters = 'affy_hg_u133a',values = UU,mart = martx)[,1])
```

```{r}
mylist=list(DU=DU_E,
            UD=UD_E,
            DD=DD_E,
            UU=UU_E)
sapply(mylist,function(x)sapply(mylist,function(y)sum(x%in%y)))
# mylist=sapply(mylist,function(x)setdiff(x,intersect(mylist$DU,mylist$DD)))
# mylist=sapply(mylist,function(x)setdiff(x,intersect(mylist$UD,mylist$UU)))
# sapply(mylist,function(x)sapply(mylist,function(y)sum(x%in%y)))
```

```{r}
du_list=getBM(attributes = c('ensembl_gene_id','hgnc_symbol','description'),filters = 'ensembl_gene_id',values = DU_E,mart = martx)
dd_list=getBM(attributes = c('ensembl_gene_id','hgnc_symbol','description'),filters = 'ensembl_gene_id',values = DD_E,mart = martx)
uu_list=getBM(attributes = c('ensembl_gene_id','hgnc_symbol','description'),filters = 'ensembl_gene_id',values = UU_E,mart = martx)
ud_list=getBM(attributes = c('ensembl_gene_id','hgnc_symbol','description'),filters = 'ensembl_gene_id',values = UD_E,mart = martx)
```

```{r}
du_list$description=sapply(strsplit(du_list$description,'[[]'),function(x)x[1])
# write_csv(du_list,path ='./data/processed/humanBrainMicroarray/AgeDown_DrugUp.csv')
du_list
```

```{r}
dd_list$description=sapply(strsplit(dd_list$description,'[[]'),function(x)x[1])
# write_csv(dd_list,path ='./data/processed/humanBrainMicroarray/AgeDown_DrugDown.csv')
dd_list
```

```{r}
uu_list$description=sapply(strsplit(uu_list$description,'[[]'),function(x)x[1])
# write_csv(uu_list,path ='./data/processed/humanBrainMicroarray/AgeUp_DrugUp.csv')
uu_list
```

```{r}
ud_list$description=sapply(strsplit(ud_list$description,'[[]'),function(x)x[1])
# write_csv(ud_list,path ='./data/processed/humanBrainMicroarray/AgeUp_DrugDown.csv')
ud_list
```

```{r}
GO2Gene=readRDS('../shared/GO2Gene/GO2GeneBP_20170925.rds')
load('../shared/GO2Gene/goterms_20170925.RData')
numbs=data.frame(t(sapply(GO2Gene,function(go){
  sapply(mylist,function(x)length(intersect(x,go)))
})))
numbs$DU2=numbs$UD+numbs$DD+numbs$UU
numbs$UD2=numbs$DU+numbs$DD+numbs$UU
numbs$DD2=numbs$UD+numbs$DU+numbs$UU
numbs$UU2=numbs$UD+numbs$DD+numbs$DU
numbs$Total=length(intersect(unique(unlist(GO2Gene)),unique(unlist(mylist))))
numbs$DUTot=length(intersect(unique(unlist(GO2Gene)),unique(mylist$DU)))
numbs$UDTot=length(intersect(unique(unlist(GO2Gene)),unique(mylist$UD)))
numbs$DDTot=length(intersect(unique(unlist(GO2Gene)),unique(mylist$DD)))
numbs$UUTot=length(intersect(unique(unlist(GO2Gene)),unique(mylist$UU)))
numbs$DU3=numbs$DUTot-numbs$DU
numbs$UD3=numbs$UDTot-numbs$UD
numbs$UU3=numbs$UUTot-numbs$UU
numbs$DD3=numbs$DDTot-numbs$DD
numbs$DU4=numbs$Total-numbs$DU-numbs$DU2-numbs$DU3
numbs$DD4=numbs$Total-numbs$DD-numbs$DD2-numbs$DD3
numbs$UD4=numbs$Total-numbs$UD-numbs$UD2-numbs$UD3
numbs$UU4=numbs$Total-numbs$UU-numbs$UU2-numbs$UU3
DUres=t(apply(numbs,1,function(x){
  fi=fisher.test(matrix(c(x['DU'],x['DU2'],x['DU3'],x['DU4']),ncol=2))
  c(DUOdds=fi$estimate,DUp=fi$p.value)
}))
numbs$DUOdds=DUres[,1]
numbs$DU.p=DUres[,2]

DDres=t(apply(numbs,1,function(x){
  fi=fisher.test(matrix(c(x['DD'],x['DD2'],x['DD3'],x['DD4']),ncol=2))
  c(DDOdds=fi$estimate,DDp=fi$p.value)
}))
numbs$DDOdds=DDres[,1]
numbs$DD.p=DDres[,2]

UUres=t(apply(numbs,1,function(x){
  fi=fisher.test(matrix(c(x['UU'],x['UU2'],x['UU3'],x['UU4']),ncol=2))
  c(UUOdds=fi$estimate,UUp=fi$p.value)
}))
numbs$UUOdds=UUres[,1]
numbs$UU.p=UUres[,2]

UDres=t(apply(numbs,1,function(x){
  fi=fisher.test(matrix(c(x['UD'],x['UD2'],x['UD3'],x['UD4']),ncol=2))
  c(UDOdds=fi$estimate,UDp=fi$p.value)
}))
numbs$UDOdds=UDres[,1]
numbs$UD.p=UDres[,2]
numbs$GOSize=numbs$DU+numbs$DU2
numbs$GOID=rownames(numbs)
numbs$GOTerm=terms[numbs$GOID]
```

```{r}
xx=numbs %>%
  dplyr::select(GOID,GOTerm,GOSize,DUOdds,DU.p,UDOdds,UD.p,UUOdds,UU.p,DDOdds,DD.p) %>%
  group_by(GOID) %>%
  mutate(max=max(DUOdds,UDOdds,DDOdds,UUOdds)) %>%
  ungroup()%>%
  arrange(desc(max))%>%
  filter(GOSize >=5 & GOSize <=50) %>%
  select(-max)
colnames(xx)=c('GOID','GOTerm','GOSize','DU','DU_p','UD','UD_p','UU','UU_p','DD','DD_p')
# write_csv(xx,'./data/processed/humanBrainMicroarray/DrugGO.csv')
xx
```

## Up in ageing & Up-regulated by pro-longevity drugs: Up regulated as response

```{r}
numbs %>%
   dplyr::select(GOID,GOTerm,GOSize,UU,UU2,UU3,UU4,UUOdds,UU.p) %>%
  filter(GOSize >=5 & GOSize <=20) %>%
  mutate(UU.padj=p.adjust(UU.p,method='fdr'))%>%
  arrange(desc(UUOdds)) %>%
filter(UUOdds >=2)
```

## Down in ageing & Up-regulated by pro-longevity drugs: Down regulation is bad, should be reversed! 

```{r}
numbs %>%
   dplyr::select(GOID,GOTerm,GOSize,DU,DU2,DU3,DU4,DUOdds,DU.p) %>%
  filter(GOSize >=5& GOSize <=20) %>%
  mutate(DU.padj=p.adjust(DU.p,method='fdr'))%>%
  arrange(desc(DUOdds)) %>%
filter(DUOdds >=2)
```

## Down in ageing & Down-regulated by pro-longevity drugs: Down regulated as response

```{r}
numbs %>%
   dplyr::select(GOID,GOTerm,GOSize,DD,DD2,DD3,DD4,DDOdds,DD.p) %>%
  filter(GOSize >=5& GOSize <=20) %>%
  mutate(DD.padj=p.adjust(DD.p,method='fdr'))%>%
  arrange(desc(DDOdds)) %>%
filter(DDOdds >=2)
```

## Up in ageing & Down-regulated by pro-longevity drugs: Up regulation is bad, should be reversed! 

```{r}
numbs %>%
   dplyr::select(GOID,GOTerm,GOSize,UD,UD2,UD3,UD4,UDOdds,UD.p) %>%
  filter(GOSize >=5& GOSize <=20) %>%
  mutate(UD.padj=p.adjust(UD.p,method='fdr'))%>%
  arrange(desc(UDOdds)) %>%
filter(UDOdds >=2)
```

## Summary of functional analysis - Ups & Downs and Causes & Responses

```{r,fig.height=16,fig.width=4}
xx=numbs %>%
  filter(GOSize>=5 & GOSize<=20)%>%
  dplyr::select(DUOdds,DDOdds,UUOdds,UDOdds,GOTerm)
colnames(xx)[1:4]=c('DownUp - Cause',
               'DownDown - Response',
               'UpUp - Response',
               'UpDown - Cause')
rownames(xx)=xx$GOTerm
xx$GOTerm=NULL
xx=log2(as.matrix(xx))
xx[xx==-Inf]=-ceiling(sort(unique(c(abs(xx))),dec=T)[2])
xx[xx==Inf]=ceiling(sort(unique(c(abs(xx))),dec=T)[2])
xx=xx[!rowSums(abs(xx)>2)==0,]
brx=seq(-max(abs(xx)),max(abs(xx)),length.out = 20)
clx=colorRampPalette(c('blue','gray95','red'))(19)
# clx[brx<1 & brx>(-1)]='gray95'
pheatmap(xx,
         color = clx,
         breaks = brx,
         cutree_cols = 2,cutree_rows = 10)
```

## The other drugs: pro- or anti-longevity?

```{r,fig.width=3,fig.height=4}
uu=combined %>%
  filter(ProbeID%in%UU) %>%
  select(ProbeID,DrugChange,Name,DrugAge) %>%
  unique() %>%
  group_by(Name,ProbeID,DrugAge) %>%
  summarise(percPos=mean(DrugChange>0))%>%
  spread(value = percPos,key = ProbeID) %>%
  as.data.frame()

ud=combined %>%
  filter(ProbeID%in%UD) %>%
  select(ProbeID,DrugChange,Name) %>%
  unique() %>%
  group_by(Name,ProbeID) %>%
  summarise(percPos=mean(DrugChange>0))%>%
  spread(value = percPos,key = ProbeID) %>%
  as.data.frame()

dd=combined %>%
  filter(ProbeID%in%DD) %>%
  select(ProbeID,DrugChange,Name) %>%
  unique() %>%
  group_by(Name,ProbeID) %>%
  summarise(percPos=mean(DrugChange>0))%>%
  spread(value = percPos,key = ProbeID) %>%
  as.data.frame()

du=combined %>%
  filter(ProbeID%in%DU) %>%
  select(ProbeID,DrugChange,Name) %>%
  unique() %>%
  group_by(Name,ProbeID) %>%
  summarise(percPos=mean(DrugChange>0))%>%
  spread(value = percPos,key = ProbeID) %>%
  as.data.frame()

drage=uu$DrugAge
uu$DrugAge=NULL
nm=uu$Name
uu$Name=NULL
ud$Name=NULL
du$Name=NULL
dd$Name=NULL
resx=data.frame(Name=nm,DrugAge=as.factor(drage),
           UU=apply(uu,1,function(x)mean(x>0.5)),
           DU=apply(du,1,function(x)mean(x>0.5)),
           DD=apply(dd,1,function(x)mean(x<0.5)),
           UD=apply(ud,1,function(x)mean(x<0.5)))
rownames(resx)=resx$Name
annotrow=(combined %>% select(Name,Array_MeanScore,DrugAge) %>% unique() %>% data.frame())
rownames(annotrow)=annotrow$Name
annotrow$Name=NULL
annotrow$DrugAge=as.factor(c('-','+')[1+(annotrow$DrugAge)])
annotrow$DrugSign=as.factor(c('-','+')[1+(annotrow$Array_MeanScore>0)])
annotrow$Array_MeanScore=NULL

# brx1=seq(-1,1,by=0.001)
# clx1=colorRampPalette(brewer.pal(3,'PrGn'))(length(brx1))
rowcolors=list(DrugSign=setNames(brewer.pal(4,'Set1')[1:2],c('+','-')),
               DrugAge=setNames(brewer.pal(4,'Set1')[3:4],c('+','-')))

brx=seq(0,1,length.out = 20)
clx=colorRampPalette(c('blue','gray95','red'))(19)
xx=resx[,3:6]
colnames(xx)=paste(colnames(xx),'_',sapply(list(uu,du,dd,ud),ncol),sep='')
pheatmap(xx,
         annotation_row =annotrow,annotation_colors = rowcolors,
         color = clx,
         breaks = brx,display_numbers = T,number_color = 'gray95',
         cutree_rows = 6)
# pheatmap(xx,
#          annotation_row =annotrow,annotation_colors = rowcolors,
#          color = clx,
#          breaks = brx,display_numbers = T,number_color = 'gray95',
#          cutree_rows = 6,filename = './results/article/drugSign.pdf')
```

```{r, fig.height=10, fig.width=15}
xx=combined%>%
  filter(ProbeID%in%c(UU,DU,DD,UD))%>%
  select(InstanceID,DrugChange,ProbeID)%>%
  unique()%>%
  spread(key = ProbeID,value = DrugChange)
inst=xx$InstanceID
xx$InstanceID=NULL
xx=as.matrix(xx)
rownames(xx)=inst

rowannot=combined %>% 
  mutate(DrugSign=Array_MeanScore>0) %>%
  select(InstanceID,Name,DrugSign,DrugAge)%>%
  unique()%>%
  data.frame()
rownames(rowannot)=rowannot$InstanceID
rowannot$InstanceID=NULL
rowannot$Name=as.factor(rowannot$Name)
rowannot$DrugAge=as.factor(c('-','+')[1+(rowannot$DrugAge)])
rowannot$DrugSign=as.factor(c('-','+')[1+(rowannot$DrugSign)])
brx=seq(-max(abs(xx)),max(abs(xx)),length.out = 100)
clx=colorRampPalette(c('blue','gray95','red'))(99)
rowcolors=list(DrugSign=setNames(brewer.pal(8,'Set1')[1:2],c('+','-')),
               DrugAge=setNames(brewer.pal(8,'Set1')[3:4],c('+','-')),
               ProLong=setNames(brewer.pal(8,'Set1')[5:6],c('Down','Up')),
               ArrayChange=setNames(brewer.pal(8,'Set1')[7:8],c('Down','Up')),
               Name=setNames(c(brewer.pal(8,'Set2'),brewer.pal(8,'Set3'),brewer.pal(8,'Dark2')),unique(rowannot$Name)))

colannot=combined %>% 
  select(ProbeID,ArrayChange) %>% 
  unique() %>%
  data.frame()

colannot$ProLong='-'
colannot$ProLong[colannot$ProbeID%in%drugPos]='Up'
colannot$ProLong[colannot$ProbeID%in%drugNeg]='Down'
rownames(colannot)=colannot$ProbeID
colannot$ProbeID=NULL
colannot=colannot[colnames(xx),]
pheatmap(xx,
         annotation_row = rowannot,annotation_colors = rowcolors,
         annotation_col = colannot,
         show_rownames = F,
         color = clx,
         breaks = brx,
         show_colnames = F,
         cutree_rows = 2, cutree_cols = 2)
```

```{r}
save(list=ls(),file='./scripts/DrugSign_gene/all.RData')
```

